Appendix 1: Evaluation of the three species in Figure 1

### 2015-10-15 Harris_2015/manuscript-materials/draft.tex
clean up directory

\documentclass[12pt,]{article}
\usepackage[compact]{titlesec}
\usepackage{fancyhdr}
\usepackage{hanging}
\pagestyle{fancy}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\usepackage{setspace}
\usepackage{geometry}
\usepackage{ifthen}
\geometry{verbose,letterpaper,tmargin=2.54cm, bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

\date{}

\begin{document}

\begin{spacing}{1.9}
\begin{flushleft}
\chead{\ifthenelse{\value{page}=1}{}{\textit{Species interactions in Markov networks}}}
\renewcommand{\headrulewidth}{0pt}

@@

To the extent that nontrophic species interactions (such as competition)
affect community assembly, ecologists might expect to find signatures of
these interactions in species composition data (MacArthur 1958, Diamond
1975). Despite decades of work and several major controversies, however
(Lewin 1983, Strong et al. 1984, Gotelli and Entsminger 2003, Connor et
al. 2013), existing methods for detecting competition's effects on
community structure are unreliable (Gotelli and Ulrich 2009). In
particular, species' effects on one another can become lost in the
complex web of direct and indirect interactions in real assemblages. For
example, the competitive interaction between the two shrub species in
Figure 1A can become obscured by their shared tendency to occur in
unshaded areas (Figure 1B). While ecologists have long known that
indirect effects can overwhelm direct ones at the landscape level
(Dodson 1970, Levine 1976), the vast majority of our methods for drawing
inferenes from observational data do not control for these effects (e.g.
Diamond 1975, Strong et al. 1984, Gotelli and Ulrich 2009, Veech 2013,
Pollock et al. 2014). To the extent that indirect interactions like
those in Figure 1 are generally important (Dodson 1970), existing
methods will thus not generally provide much evidence regarding species'
direct effects on one another. The goal of this paper is to resolve this
long-standing problem.

While competition doesn't reliably reduce co-occurrence rates at the
whole-landscape level (as most of our methods assume), it nevertheless
does leave a signal in the data (Figure 1C). Specifically, after
partitioning the data set into shaded sites and unshaded sites, there
will be co-occurrence deficits in each subset that might not be apparent
at the landscape level. More generally, we can obtain much better
estimates of the association between two species from their conditional
relationships (i.e.~by controlling for other species in the network)
than we could get from their overall co-occurrence rates. This kind of
precision is difficult to obtain from null models, which begin with the
assumption that all the pairwise interactions are zero and thus don't
need to be controlled for. Nevertheless, null models have dominated this
field for more than three decades (Strong et al. 1984, Gotelli and
Ulrich 2009).

Following recent work by Azaele et al. (2010) and Fort (2013), this
paper shows that Markov networks (undirected graphical models also known
as Markov random fields; Murphy 2012) can provide a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observed presence-absence
matrices. Markov networks have been used in many scientific fields in
similar contexts for decades, from physics (where nearby particles
interact magnetically; Cipra 1987) to spatial statistics (where adjacent
grid cells have correlated values; Harris 1974, Gelfand et al. 2005).
While community ecologists explored some related approaches in the
1980's (Whittam and Siegel-Causey 1981), they used severe approximations
that led to unintelligible results (e.g. ``probabilities'' greater than
one; Gilpin and Diamond 1982).

Below, I introduce Markov networks and show how they can be used to
simulate landscape-level data or to make exact predictions about the
direct and indirect consequences of possible interaction matrices. Then,
using simulated data sets where the ``true'' interactions are known, I
compare this approach with several existing methods. Finally, I discuss
opportunities for extending the approach presented here to other
problems in community ecology, e.g.~quantifying the overall effect of
species interactions on occurrence rates (Roughgarden 1983) and
disentangling the effects of biotic versus abiotic interactions on
species composition (Kissling et al. 2012, Pollock et al. 2014).

@@

Of course, if all else is \emph{not} equal (e.g.~Figure 1, where the
presence of one competitor is associated with release from another
competitor), then species' marginal association rates can differ from
this expectation. Determining the marginal relationships between species
from their conditional interactions entails summing over the different
possible assemblages (Figure 2B). This becomes intractable when the
number of possible assemblages is large, though several methods beyond
the scope of this paper can be employed to keep the calculations
feasible (Salakhutdinov 2008, Lee and Hastie 2012). Alternatively, as
noted below, some common linear and generalized linear methods can also
be used as computationally efficient approximations to the full network
(Lee and Hastie 2012, Loh and Wainwright 2013).

@@

In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a matrix of ones and zeros indicating which species are present at which
sites). When the number of species is reasonably small, one can compute
exact maximum likelihood estimates for all of the \(\alpha\) and
\(\beta\) coefficients given a presence-absence matrix by optimizing
\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces (Murphy 2012),
ensuring that this procedure will always converge on the global maximum.
This maximum represents the unique combination of \(\alpha\) and
\(\beta\) coefficients that would be expected to produce exactly the
observed co-occurrence frequencies on average (i.e.~maximizing the
likelihood matches the sufficient statistics of the model distribution
to the sufficient statistics of the data; Murphy 2012). I used the
rosalia package (Harris 2015a) for the R programming language (R Core
Team 2015) to optimize the Markov network parameters. The package was
named after Santa Rosalia, the patron saint of biodiversity, whose
supposedly miraculous healing powers played an important rhetorical role
in the null model debates of the 1970's and 1980's (Lewin 1983).

@@

I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package (Harris 2015a) to fit Markov newtork
models, as described above. For the analyses with 20 species, I added a
very weak logistic prior distribution on the \(\alpha\) and \(\beta\)
terms with scale 2 to ensure that the model estimates were always
finite. The bias introduced by this prior should be small: the 95\%
credible interval on \(\beta\) only requires that one species' effect on
the odds of observing a different species to be less than a factor of
1500 (which is not much of a constraint). The logistic distribution was
chosen because it is convex and has a similar shape to the Laplace
distribution used in LASSO regularization (especially in the tails), but
unlike the Laplace distribution it is differentiable everywhere and does
not force any estimates to be exactly zero. To confirm that this
procedure produced stable estimates, I compared its estimates on 50
bootstrap replicates (Appendix 4).

I also evaluated six alternative methods: five from the existing
literature, plus a novel combination of two of these methods. The first
alternative interaction metric was the sample correlation between
species' presence-absence vectors, which summarizes their marginal
association. Next, I used partial correlations, which summarize species'
conditional relationships (Albrecht and Gotelli 2001, Faisal et al.
2010). In the context of non-Gaussian data, the partial correlation can
be thought of as a computationally efficient approximation to the full
Markov network model (Loh and Wainwright 2013). This sort of model is
very common for estimating relationships among genes and gene products
(Friedman et al. 2008). Because partial correlations are undefined for
landscapes with perfectly-correlated species pairs, I used a regularized
estimate based on James-Stein shrinkage, as implemented in the corpcor
package's \texttt{pcor.shrink} function with the default settings
(Schäfer et al. 2014).

The third alternative, generalized linear models (GLMs), can also be
thought of as a computationally efficient approximation to the Markov
network (Lee and Hastie 2012). Following Faisal et al. (2010), I fit
regularized logistic regression models (Gelman et al. 2008) for each
species, using the other species on the landscape as predictors. To
avoid the identifiability problems associated with directed cyclic
graphs (Schmidt and Murphy 2012), I then symmetrized the relationships
within species pairs via averaging.

The next method, described in Gotelli and Ulrich (2009), involved
simulating new landscapes from a null model that retains the row and
column sums of the original matrix (Strong et al. 1984). I used the
\(Z\)-scores computed by the Pairs software described in Gotelli and
Ulrich (2009) as my null model-based estimator of species interactions.

The last two estimators used the latent correlation matrix estimated by
the BayesComm package (Golding and Harris 2015) in order to evaluate the
recent claim that the correlation coefficients estimated by ``joint
species distribution models'' provide an accurate assessment of species'
pairwise interactions (Pollock et al. 2014, see also Harris 2015b). In
addition to using the posterior mean correlation (Pollock et al. 2014),
I also used the posterior mean \emph{partial} correlation, which might
be able to control for indirect effects.

@@

Despite some variability across contexts (Figure 3A), the four methods
that controlled for indirect effects clearly performed the best: the
Markov network explained the largest portion of the variance in the
``true'' interaction coefficients (35\% overall), followed by the
generalized linear models (30\%), partial correlations from the raw
presence-absence data (28\%), and partial correlations from BayesComm,
the joint species distribution model (26\%). The benefit of choosing the
full Markov network over the other three methods was largest on the
smaller landscapes, which are also the ones that are most representative
of typical analyses in this field (Gotelli and Ulrich 2009).

@@

The results presented above show that Markov networks can reliably
recover species' pairwise interactions from observational data, even for
cases where a common null modeling technique reliably fails.
Specifically, Markov networks were successful even when direct
interactions were largely overwhelmed by indirect effects (Figure 1).
For cases where fitting a Markov network is computationally infeasible,
these results also indicate that partial covariances and generalized
linear models (the two methods that estimated conditional relationships
rather than marginal ones) can both provide useful approximations. The
partial correlations' success on simulated data may not carry over to
real data sets, however; Loh and Wainwright (2013) show that the linear
approximations can be less reliable in cases where the true interaction
matrix contains more structure (e.g.~guilds or trophic levels).
Similarly, the approximation involved in using separate generalized
linear models for each species can occasionally lead to catastrophic
overfitting with small-to-moderate sample sizes (Lee and Hastie 2012).
For these reasons, it will usually be best to fit a Markov network
rather than one of the alternative methods when one's computational
resources allow it.

It's important to note that none of these methods can identify the exact
nature of the pairwise interactions (e.g.~which species in a
positively-associated pair is facilitating the other; Schmidt and Murphy
2012), particularly when real pairs of species can reciprocally
influence one another in multiple ways simultaneously (Bruno et al.
2003); with compositional data, there is only enough information to
provide a single number describing each species pair. To estimate
asymmetric interactions, such as commensalism or predation, ecologists
would need other kinds of data, as from time series, behavioral
observations, manipulative experiments, or natural history. These other
sources of information could also be used to augment the likelihood
function with an informative prior distribution, which could lead to
better results on some real data sets than was shown in Figure 3A.

Despite their limitations, Markov networks have enormous potential to
improve ecological understanding. In particular, they are less
vulnerable than some of the most commonly-used methods to mistakenly
identifying positive species interactions between competing species, and
can make precise statements about the conditions where indirect
interactions will overwhelm direct ones. They also provide a simple
answer to the question of how competition should affect a species'
overall prevalence, which was a major flashpoint for the null model
debates in the 1980's (Roughgarden 1983, Strong et al. 1984). Equation 1
can be used to calculate the expected prevalence of a species in the
absence of biotic influences (\(\frac{e^\alpha}{1 + e^{\alpha}}\); Lee
and Hastie 2012). Competition's effect on prevalence in a Markov network
can then be calculated by subtracting this value from the observed
prevalence (cf Figure 2D). This kind of insight would have been
difficult to obtain without a generative model that makes predictions
about the consequences of species interactions; null models (which
presume \emph{a priori} that interactions do not exist) have no way to
make such predictions.

Markov networks---particularly the Ising model for binary
networks---have been studied for nearly a century (Cipra 1987), and the
models' properties, capabilities, and limits are well-understood in a
huge range of applications. Using the same framework for species
interactions would thus allow ecologists to tap into an enormous set of
existing discoveries and techniques for dealing with indirect effects,
stability, and alternative stable states. Numerous other extensions are
possible: for example, the states of the interaction network can be
modeled as a function of the local abiotic environment (Lee and Hastie
2012), which would provide a rigorous and straightforward approach to
the difficult and important task of incorporating whole networks of
biotic interactions into species distribution models (Kissling et al.
2012, Pollock et al. 2014), leading to a better understanding of the
interplay between biotic and abiotic effects on community structure.
There are even methods (Whittam and Siegel-Causey 1981, Tjelmeland and
Besag 1998) that would allow one species to affect the sign or strength
of the relationship between two other species, tipping the balance
between facilitation and exploitation (Bruno et al. 2003).

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species
interactions. Null and neutral models can be useful for clarifying our
thinking about the numerical consequences of species' richness and
abundance patterns (Harris et al. 2011, Xiao et al. 2015), but
deviations from a particular null model must be interpreted with care
(Roughgarden 1983). Even in small networks with three species, it may
simply not be possible to implicate individual species pairs or specific
ecological processes like competition by rejecting a general-purpose
null (Gotelli and Ulrich 2009), especially when the test statistic is
effectively just a correlation coefficient (Figure 3B). Simultaneous
estimation of multiple ecological parameters seems like a much more
promising approach: to the extent that the models' relative performance
on real data sets is similar to the range of results shown in Figure 3A,
scientists in this field could often double their explanatory power by
switching from null models to Markov networks (or increase it
substantially with linear or generalized linear approximations).
Regardless of the methods ecologists ultimately choose, controlling for
indirect effects could clearly improve our understanding of species'
direct effects on one another and on community structure.

@@

Albrecht, M., and N. J. Gotelli. 2001. Spatial and temporal niche
partitioning in grassland ants. Oecologia 126:134--141.

Azaele, S., R. Muneepeerakul, A. Rinaldo, and I. Rodriguez-Iturbe. 2010.
Inferring plant ecosystem organization from species occurrences. Journal
of theoretical biology 262:323--329.

Bruno, J. F., J. J. Stachowicz, and M. D. Bertness. 2003. Inclusion of
facilitation into ecological theory. Trends in Ecology \& Evolution
18:119--125.

Cipra, B. A. 1987. An introduction to the Ising model. American
Mathematical Monthly 94:937--959.

Connor, E. F., M. D. Collins, and D. Simberloff. 2013. The checkered
history of checkerboard distributions. Ecology 94:2403--2414.

Diamond, J. M. 1975. The island dilemma: Lessons of modern biogeographic
studies for the design of natural reserves. Biological conservation
7:129--146.

Dodson, S. I. 1970. COMPLEMENTARY FEEDING NICHES SUSTAINED BY
SIZE-SELECTIVE PREDATION. Limnology and Oceanography 15:131--137.

Faisal, A., F. Dondelinger, D. Husmeier, and C. M. Beale. 2010.
Inferring species interaction networks from species abundance data: A
comparative evaluation of various statistical and machine learning
methods. Ecological Informatics 5:451--464.

Fort, H. 2013. Statistical Mechanics Ideas and Techniques Applied to
Selected Problems in Ecology. Entropy 15:5237--5276.

Friedman, J., T. Hastie, and R. Tibshirani. 2008. Sparse inverse
covariance estimation with the graphical lasso. Biostatistics
9:432--441.

Gelfand, A. E., A. M. Schmidt, S. Wu, J. A. Silander, A. Latimer, and A.
G. Rebelo. 2005. Modelling species diversity through species level
hierarchical modelling. Journal of the Royal Statistical Society: Series
C (Applied Statistics) 54:1--20.

Gelman, A., A. Jakulin, M. G. Pittau, and Y.-S. Su. 2008. A Weakly
Informative Default Prior Distribution for Logistic and Other Regression
Models. The Annals of Applied Statistics 2:1360--1383.

Gilpin, M. E., and J. M. Diamond. 1982. Factors contributing to
non-randomness in species Co-occurrences on Islands. Oecologia
52:75--84.

Golding, N., and D. J. Harris. 2015. BayesComm: Bayesian Community
Ecology Analysis.

Gotelli, N. J., and G. L. Entsminger. 2003. Swap algorithms in null
model analysis. Ecology:532--535.

Gotelli, N. J., and W. Ulrich. 2009. The empirical Bayes approach as a
tool to identify non-random species associations. Oecologia
162:463--477.

Harris, D. J. 2015a. Rosalia: Exact inference for small binary Markov
networks. R package version 0.1.0. Zenodo.
http://dx.doi.org/10.5281/zenodo.17808.

Harris, D. J. 2015b. Generating realistic assemblages with a Joint
Species Distribution Model. Methods in Ecology and Evolution.

Harris, D. J., K. G. Smith, and P. J. Hanly. 2011. Occupancy is
nine-tenths of the law: Occupancy rates determine the homogenizing and
differentiating effects of exotic species. The American naturalist
177:535.

Harris, T. E. 1974. Contact Interactions on a Lattice. The Annals of
Probability 2:969--988.

Kissling, W. D., C. F. Dormann, J. Groeneveld, T. Hickler, I. Kühn, G.
J. McInerny, J. M. Montoya, C. Römermann, K. Schiffers, F. M. Schurr, A.
Singer, J.-C. Svenning, N. E. Zimmermann, and R. B. O'Hara. 2012.
Towards novel approaches to modelling biotic interactions in
multispecies assemblages at large spatial extents. Journal of
Biogeography 39:2163--2178.

Lee, J. D., and T. J. Hastie. 2012, May. Learning Mixed Graphical
Models.

Levine, S. H. 1976. Competitive Interactions in Ecosystems. The American
Naturalist 110:903--910.

Lewin, R. 1983. Santa Rosalia Was a Goat. Science 221:636--639.

Loh, P.-L., and M. J. Wainwright. 2013. Structure estimation for
discrete graphical models: Generalized covariance matrices and their
inverses. The Annals of Statistics 41:3022--3049.

MacArthur, R. H. 1958. Population ecology of some warblers of
northeastern coniferous forests. Ecology 39:599--619.

Murphy, K. P. 2012. Machine Learning: A Probabilistic Perspective. The
MIT Press.

Pollock, L. J., R. Tingley, W. K. Morris, N. Golding, R. B. O'Hara, K.
M. Parris, P. A. Vesk, and M. A. McCarthy. 2014. Understanding
co-occurrence by modelling species simultaneously with a Joint Species
Distribution Model (JSDM). Methods in Ecology and Evolution:n/a--n/a.

R Core Team. 2015. R: A Language and Environment for Statistical
Computing. R Foundation for Statistical Computing, Vienna, Austria.

Roughgarden, J. 1983. Competition and Theory in Community Ecology. The
American Naturalist 122:583--601.

Salakhutdinov, R. 2008. Learning and evaluating Boltzmann machines.
Technical Report UTML TR 2008-002, Department of Computer Science,
University of Toronto, Dept. of Computer Science, University of Toronto.

>>>>

\chead{\ifthenelse{\value{page}=1}{}{\textit{Species interactions in Markov networks}}}
\renewcommand{\headrulewidth}{0pt}

@@

To the extent that nontrophic species interactions (such as competition)
affect community assembly, ecologists might expect to find signatures of
these interactions in species composition data
\citep{macarthur_population_1958, diamond_island_1975}. Despite decades
of work and several major controversies, however
\citep{lewin_santa_1983, strong_ecological_1984, gotelli_swap_2003, connor_checkered_2013},
existing methods for detecting competition's effects on community
structure are unreliable \citep{gotelli_empirical_2009}. In particular,
species' effects on one another can become lost in the complex web of
direct and indirect interactions in real assemblages. For example, the
competitive interaction between the two shrub species in Figure 1A can
become obscured by their shared tendency to occur in unshaded areas
(Figure 1B). While ecologists have long known that indirect effects can
overwhelm direct ones at the landscape level
\citep{dodson_complementary_1970, levine_competitive_1976}, the vast
majority of our methods for drawing inferenes from observational data do
not control for these effects
\citep[e.g.][]{diamond_island_1975, strong_ecological_1984, gotelli_empirical_2009, veech_probabilistic_2013, pollock_understanding_2014}.
To the extent that indirect interactions like those in Figure 1 are
generally important \citep{dodson_complementary_1970}, existing methods
will thus not generally provide much evidence regarding species' direct
effects on one another. The goal of this paper is to resolve this
long-standing problem.

\begin{figure}[htbp]
\centering
\includegraphics{figures/Figure_1.pdf}
\caption{\textbf{Figure 1.} \textbf{A.} A small network of three
competing species. The tree (top) tends not to co-occur with either of
the two shrub species, as indicated by the strongly negative coefficient
linking them. The two shrub species also compete with one another, but
more weakly (circled coefficient). \textbf{B.} In spite of the
competitive interactions between the two shrub species, their shared
tendency to occur in locations without trees makes their occurrence
vectors positively correlated (circled). \textbf{C.} Controlling for the
tree species' presence with a conditional method such as a partial
covariance or a Markov network leads to correct identification of the
negative shrub-shrub interaction (circled).}
\end{figure}

While competition doesn't reliably reduce co-occurrence rates at the
whole-landscape level (as most of our methods assume), it nevertheless
does leave a signal in the data (Figure 1C). Specifically, after
partitioning the data set into shaded sites and unshaded sites, there
will be co-occurrence deficits in each subset that might not be apparent
at the landscape level. More generally, we can obtain much better
estimates of the association between two species from their conditional
relationships (i.e.~by controlling for other species in the network)
than we could get from their overall co-occurrence rates. This kind of
precision is difficult to obtain from null models, which begin with the
assumption that all the pairwise interactions are zero and thus don't
need to be controlled for. Nevertheless, null models have dominated this
field for more than three decades
\citep{strong_ecological_1984, gotelli_empirical_2009}.

Following recent work by \citet{azaele_inferring_2010} and
\citet{fort_statistical_2013}, this paper shows that Markov networks
\citep[undirected graphical models also known as Markov random
fields;][]{murphy_machine_2012} can provide a framework for
understanding the landscape-level consequences of pairwise species
interactions, and for detecting them from observed presence-absence
matrices. Markov networks have been used in many scientific fields in
similar contexts for decades, from physics \citep[where nearby particles
interact magnetically;][]{cipra_introduction_1987} to spatial statistics
\citep[where adjacent grid cells have correlated
values;][]{harris_contact_1974, gelfand_modelling_2005}. While community
ecologists explored some related approaches in the 1980's
\citep{whittam_species_1981}, they used severe approximations that led
to unintelligible results \citep[e.g. ``probabilities'' greater than
one;][]{gilpin_factors_1982}.

Below, I introduce Markov networks and show how they can be used to
simulate landscape-level data or to make exact predictions about the
direct and indirect consequences of possible interaction matrices. Then,
using simulated data sets where the ``true'' interactions are known, I
compare this approach with several existing methods. Finally, I discuss
opportunities for extending the approach presented here to other
problems in community ecology, e.g.~quantifying the overall effect of
species interactions on occurrence rates
\citep{roughgarden_competition_1983} and disentangling the effects of
biotic versus abiotic interactions on species composition
\citep{kissling_towards_2012, pollock_understanding_2014}.

@@

\begin{figure}[htbp]
\centering
\includegraphics{figures/Figure_2.pdf}
\caption{\textbf{Figure 2.} \textbf{A.} A small Markov network with two
species. The abiotic environment favors the occurrence of both species
(\(\alpha >0\)), particularly species 2 (\(\alpha_2 > \alpha_1\)). The
negative \(\beta\) coefficient linking these two species implies that
they co-occur less than expected under independence. \textbf{B.}
Relative probabilities of all four possible presence-absence
combinations for Species 1 and Species 2. The exponent includes
\(\alpha_1\) whenever Species 1 is present (\(y_1 = 1\)), but not when
it is absent (\(y_1 = 0\)). Similarly, the exponent includes
\(\alpha_2\) only when species \(2\) is present (\(y_2 = 1\)), and
\(\beta\) only when both are present (\(y_1y_2 = 1\)). The normalizing
constant \(Z\), ensures that the four relative probabilities sum to 1.
In this case, \(Z\) is about 18.5. \textbf{C.} We can find the expected
frequencies of all possible co-occurrence patterns between the two
species of interest. \textbf{D.} If \(\beta_{12}\) equaled zero (e.g.~if
the species no longer competed for the same resources), then the
reduction in competition would allow each species to increase its
occurrence rate and the co-occurrence deficit would be eliminated.}
\end{figure}

Of course, if all else is \emph{not} equal (e.g.~Figure 1, where the
presence of one competitor is associated with release from another
competitor), then species' marginal association rates can differ from
this expectation. Determining the marginal relationships between species
from their conditional interactions entails summing over the different
possible assemblages (Figure 2B). This becomes intractable when the
number of possible assemblages is large, though several methods beyond
the scope of this paper can be employed to keep the calculations
feasible \citep{lee_learning_2012, salakhutdinov_learning_2008}.
Alternatively, as noted below, some common linear and generalized linear
methods can also be used as computationally efficient approximations to
the full network \citep{lee_learning_2012, loh_structure_2013}.

@@

In the previous section, the values of \(\alpha\) and \(\beta\) were
known and the goal was to make predictions about possible species
assemblages. In practice, however, ecologists will often need to
estimate the parameters from an observed co-occurrence matrix (i.e.~from
a matrix of ones and zeros indicating which species are present at which
sites). When the number of species is reasonably small, one can compute
exact maximum likelihood estimates for all of the \(\alpha\) and
\(\beta\) coefficients given a presence-absence matrix by optimizing
\(p(\vec{y}; \alpha, \beta)\). Fully-observed Markov networks like the
ones considered here have unimodal likelihood surfaces
\citep{murphy_machine_2012}, ensuring that this procedure will always
converge on the global maximum. This maximum represents the unique
combination of \(\alpha\) and \(\beta\) coefficients that would be
expected to produce exactly the observed co-occurrence frequencies on
average \citep[i.e.~maximizing the likelihood matches the sufficient
statistics of the model distribution to the sufficient statistics of the
data;][]{murphy_machine_2012}. I used the rosalia package
\citep{harris_rosalia_2015} for the R programming language
\citep{r_core_team_r_2015} to optimize the Markov network parameters.
The package was named after Santa Rosalia, the patron saint of
biodiversity, whose supposedly miraculous healing powers played an
important rhetorical role in the null model debates of the 1970's and
1980's \citep{lewin_santa_1983}.

@@

I compared seven techniques for determining the sign and strength of the
associations between pairs of species from simulated data (Appendix 3).
First, I used the rosalia package \citep{harris_rosalia_2015} to fit
Markov newtork models, as described above. For the analyses with 20
species, I added a very weak logistic prior distribution on the
\(\alpha\) and \(\beta\) terms with scale 2 to ensure that the model
estimates were always finite. The bias introduced by this prior should
be small: the 95\% credible interval on \(\beta\) only requires that one
species' effect on the odds of observing a different species to be less
than a factor of 1500 (which is not much of a constraint). The logistic
distribution was chosen because it is convex and has a similar shape to
the Laplace distribution used in LASSO regularization (especially in the
tails), but unlike the Laplace distribution it is differentiable
everywhere and does not force any estimates to be exactly zero. To
confirm that this procedure produced stable estimates, I compared its
estimates on 50 bootstrap replicates (Appendix 4).

I also evaluated six alternative methods: five from the existing
literature, plus a novel combination of two of these methods. The first
alternative interaction metric was the sample correlation between
species' presence-absence vectors, which summarizes their marginal
association. Next, I used partial correlations, which summarize species'
conditional relationships
\citep{albrecht_spatial_2001, faisal_inferring_2010}. In the context of
non-Gaussian data, the partial correlation can be thought of as a
computationally efficient approximation to the full Markov network model
\citep{loh_structure_2013}. This sort of model is very common for
estimating relationships among genes and gene products
\citep{friedman_sparse_2008}. Because partial correlations are undefined
for landscapes with perfectly-correlated species pairs, I used a
regularized estimate based on James-Stein shrinkage, as implemented in
the corpcor package's \texttt{pcor.shrink} function with the default
settings \citep{schafer_corpcor_2014}.

The third alternative, generalized linear models (GLMs), can also be
thought of as a computationally efficient approximation to the Markov
network \citep{lee_learning_2012}. Following
\citet{faisal_inferring_2010}, I fit regularized logistic regression
models \citep{gelman_weakly_2008} for each species, using the other
species on the landscape as predictors. To avoid the identifiability
problems associated with directed cyclic graphs
\citep{schmidt_modeling_2012}, I then symmetrized the relationships
within species pairs via averaging.

The next method, described in \citet{gotelli_empirical_2009}, involved
simulating new landscapes from a null model that retains the row and
column sums of the original matrix \citep{strong_ecological_1984}. I
used the \(Z\)-scores computed by the Pairs software described in
\citet{gotelli_empirical_2009} as my null model-based estimator of
species interactions.

The last two estimators used the latent correlation matrix estimated by
the BayesComm package \citep{golding_bayescomm_2015} in order to
evaluate the recent claim that the correlation coefficients estimated by
``joint species distribution models'' provide an accurate assessment of
species' pairwise interactions
\citetext{\citealp{pollock_understanding_2014}; \citealp[see
also][]{harris_generating_2015}}. In addition to using the posterior
mean correlation \citep{pollock_understanding_2014}, I also used the
posterior mean \emph{partial} correlation, which might be able to
control for indirect effects.

@@

Despite some variability across contexts (Figure 3A), the four methods
that controlled for indirect effects clearly performed the best: the
Markov network explained the largest portion of the variance in the
``true'' interaction coefficients (35\% overall), followed by the
generalized linear models (30\%), partial correlations from the raw
presence-absence data (28\%), and partial correlations from BayesComm,
the joint species distribution model (26\%). The benefit of choosing the
full Markov network over the other three methods was largest on the
smaller landscapes, which are also the ones that are most representative
of typical analyses in this field \citep{gotelli_empirical_2009}.

\begin{figure}[htbp]
\centering
\includegraphics{figures/performance.pdf}
\caption{\textbf{Figure 3.} \textbf{A.} Proportion of variance in
interaction coefficients explained by each method versus number of
sampled locations. \textbf{B.} The \(Z\)-scores produced by the null
model (``Pairs'') for each pair of species can be predicted using the
correlation between the presence-absence vectors of those same species
and from the number of sites on the landscape.}
\end{figure}

@@

The results presented above show that Markov networks can reliably
recover species' pairwise interactions from observational data, even for
cases where a common null modeling technique reliably fails.
Specifically, Markov networks were successful even when direct
interactions were largely overwhelmed by indirect effects (Figure 1).
For cases where fitting a Markov network is computationally infeasible,
these results also indicate that partial covariances and generalized
linear models (the two methods that estimated conditional relationships
rather than marginal ones) can both provide useful approximations. The
partial correlations' success on simulated data may not carry over to
real data sets, however; \citet{loh_structure_2013} show that the linear
approximations can be less reliable in cases where the true interaction
matrix contains more structure (e.g.~guilds or trophic levels).
Similarly, the approximation involved in using separate generalized
linear models for each species can occasionally lead to catastrophic
overfitting with small-to-moderate sample sizes
\citep{lee_learning_2012}. For these reasons, it will usually be best to
fit a Markov network rather than one of the alternative methods when
one's computational resources allow it.

It's important to note that none of these methods can identify the exact
nature of the pairwise interactions \citep[e.g.~which species in a
positively-associated pair is facilitating the
other;][]{schmidt_modeling_2012}, particularly when real pairs of
species can reciprocally influence one another in multiple ways
simultaneously \citep{bruno_inclusion_2003}; with compositional data,
there is only enough information to provide a single number describing
each species pair. To estimate asymmetric interactions, such as
commensalism or predation, ecologists would need other kinds of data, as
from time series, behavioral observations, manipulative experiments, or
natural history. These other sources of information could also be used
to augment the likelihood function with an informative prior
distribution, which could lead to better results on some real data sets
than was shown in Figure 3A.

Despite their limitations, Markov networks have enormous potential to
improve ecological understanding. In particular, they are less
vulnerable than some of the most commonly-used methods to mistakenly
identifying positive species interactions between competing species, and
can make precise statements about the conditions where indirect
interactions will overwhelm direct ones. They also provide a simple
answer to the question of how competition should affect a species'
overall prevalence, which was a major flashpoint for the null model
debates in the 1980's
\citep{roughgarden_competition_1983, strong_ecological_1984}. Equation 1
can be used to calculate the expected prevalence of a species in the
absence of biotic influences
\citep[\(\frac{e^\alpha}{1 + e^{\alpha}}\);][]{lee_learning_2012}.
Competition's effect on prevalence in a Markov network can then be
calculated by subtracting this value from the observed prevalence (cf
Figure 2D). This kind of insight would have been difficult to obtain
without a generative model that makes predictions about the consequences
of species interactions; null models (which presume \emph{a priori} that
interactions do not exist) have no way to make such predictions.

Markov networks---particularly the Ising model for binary
networks---have been studied for nearly a century
\citep{cipra_introduction_1987}, and the models' properties,
capabilities, and limits are well-understood in a huge range of
applications. Using the same framework for species interactions would
thus allow ecologists to tap into an enormous set of existing
discoveries and techniques for dealing with indirect effects, stability,
and alternative stable states. Numerous other extensions are possible:
for example, the states of the interaction network can be modeled as a
function of the local abiotic environment \citep{lee_learning_2012},
which would provide a rigorous and straightforward approach to the
difficult and important task of incorporating whole networks of biotic
interactions into species distribution models
\citep{kissling_towards_2012, pollock_understanding_2014}, leading to a
better understanding of the interplay between biotic and abiotic effects
on community structure. There are even methods
\citep{whittam_species_1981, tjelmeland_markov_1998} that would allow
one species to affect the sign or strength of the relationship between
two other species, tipping the balance between facilitation and
exploitation \citep{bruno_inclusion_2003}.

Finally, the results presented here have important implications for
ecologists' continued use of null models for studying species
interactions. Null and neutral models can be useful for clarifying our
thinking about the numerical consequences of species' richness and
abundance patterns \citep{harris_occupancy_2011, xiao_strong_2015}, but
deviations from a particular null model must be interpreted with care
\citep{roughgarden_competition_1983}. Even in small networks with three
species, it may simply not be possible to implicate individual species
pairs or specific ecological processes like competition by rejecting a
general-purpose null \citep{gotelli_empirical_2009}, especially when the
test statistic is effectively just a correlation coefficient (Figure
3B). Simultaneous estimation of multiple ecological parameters seems
like a much more promising approach: to the extent that the models'
relative performance on real data sets is similar to the range of
results shown in Figure 3A, scientists in this field could often double
their explanatory power by switching from null models to Markov networks
(or increase it substantially with linear or generalized linear
approximations). Regardless of the methods ecologists ultimately choose,
controlling for indirect effects could clearly improve our understanding
of species' direct effects on one another and on community structure.

@@
